Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-26 submission/manuscript.Rmd
formatting edits WGBS figure
results WGBS edits

## KDM5C promotes _de novo_ DNA methylation at germline genes harboring CpG islands

@@

Based on the above observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs. To test this hypothesis, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3) in our previously published ChIP-seq datasets of the wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022]. In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 6A). We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B). We next evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation. We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation into EpiLCs (Figure 6C). While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6E).

@@

In wild-type cells, germline genes are known to accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012]. We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023] and found out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) contained CGIs within their promoters (TSS &#177; 500 bp) (Figure 6F). CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) (Figure 6G).

To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs). We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during ESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp). The majority of germline genes gained substantial CpGme at their promoter, regardless if their promoter contained a CGI (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%) (Figure 6H).

We then assessed differentially methylated promoters in wild-type versus _Kdm5c_-KO exEpiLCs and found `r nrow(WTvsKO_hypo)` germline gene promtoers were significantly hypomethylated with _Kdm5c_ loss (Figure 6I). Hypomethylated promoters included genes consistently dysregulated in _Kdm5c_-KO cells, such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6J).

- Then compared CpGme accumulation in wild-type vs Kdm5c-KO exEpiLCs. 
	- Out of the CGI genes, which had significantly reduced CpGme in 5CKO (heatmap of % methylation)
	- Highlight interesting genes affected vs unaffected by KDM5C (especially if same/different from E2F6, PRC1.6, Setdb1 targets)
	- Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 6F). 
		- CGI vs Non-CGI germline genes, any significant differences?

- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via DNA methylation at CGIs during early embryogenesis.

>>>>

## KDM5C is recruited to germline gene promoters harboring CpG islands to facilitate _de novo_ DNA methylation

@@

Based on the above observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs. To test this hypothesis, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3), at germline gene promoters in our previously published ChIP-seq datasets in male wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022]. In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 6A). We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B). To elucidate KDM5C's embyonic role, we then characterized KDM5C's mRNA and protein expression during male ESC to EpiLC differentiation (Figure 6C). While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6E). Together, these data suggest KDM5C acts during the transition between ESCs and EpiLCs to remove H3K4me at germline gene promoters.

@@

In wild-type cells, germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012], reaching peak methylation levels when differentiated into EpiLCs for 96 hours (extended EpiLCs, exEpiLCs)[@mochizukiRepressionGermlineGenes2021]. We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023] and found out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) contained CGIs within their promoters (TSS &#177; 500 bp) (Figure 6F). CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) when compared to CGI-free genes (Figure 6G). Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline gene promoters (FISHER'S XXXX), with 79% of KDM5C-bound germline genes containing CGIs (Figure 6G).

We then perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) to assess how KDM5C loss impacted initial CpGme placement at germline gene promoters (Figure 6H). We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during ESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp). In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6I).

We then identified germline gene promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp) and found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6J). Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated DMRs). We then evaluated promoter CpGme at germline genes ectopically transcibed in either _Kdm5c_-KO EpiLCs or within the brain and found promoter CpGme was substantially reduced in about half of germline DEGs (Figure 6K). Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets, such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L). Suprisingly, we only observed a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M). Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at select germline gene promoters during early embryogenesis, however other germline gene silencing mechansism can sufficiently compensate for KDM5C's loss at select germline gene promoters.
